Technique for \earlywarning\ detection of compact binary coalescence and its implication for multi-messenger astronomy

### 2011-06-03 method.tex
Updated flops section
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@861 ccdd05d5-56c1-461a-bf98-85f6876523b4

We now examine the computational cost scaling of the approximate implementation
of \eqref{eq:SNRTD} as \eqref{eq:decomp}.  An actual implementation of this
decomposition in a working analysis pipeline is discussed in the next section
along with measured computational requirements.  For convenience, table
\ref{tab:recap} is a recap of the meaning of various symbols used in this
calculation.
%
% symbols used in FLOPs calculations table
%
\begin{table}
\begin{tabular}{rl}
\bf{Symbol}	& \bf{Definition} \\
\hline
\tmpsamps	& The number of sample points per template \\
\numtmps		& The number of templates \\
\numslices	& The number of time slices \\
\numsvdtmps	& The number of orthogonal templates in time slice $s$ \\
\slicessamps	& The number of samples in time slice $s$ \\
$f^s$		& The sample rate in time slice $s$ \\
\resampsamps	& The number of coefficients in the decimation and interpolation filters
\end{tabular}
\caption{\label{tab:recap} Notation used to describe filters.  This table
provides a quick reference for symbols used.}
\end{table}

In table \ref{table:flops} we present the computational cost scaling in
floating point operations per sample for common tasks in the pipeline.  
%
% common tasks FLOPS table
%
\begin{table}[htdp]
\begin{center}
\begin{tabular}{l l}
\bf{Process} & \bf{ops/sample} \\
\hline
% FIR
\fir\ matched filter, \numtmps\ templates of length \tmpsamps\ & $2 \numtmps \tmpsamps$ \\
% FFT
\fft\ matched filter, \numtmps\ templates of length \tmpsamps\, & \multirow{2}{*}{$\mbox{\fontsize{14}{17}\selectfont $\frac{4 (\numtmps + 1) \lg \fftblock + 2 \numtmps}{1 - \tmpsamps/\fftblock}$}$} \\
% FFT contd...
$\,\,\,\,\,\,$ blocks of length \fftblock & \\
% Resampling
\fir\ resampling filter, length \resampsamps\ for each of \numtmps\ templates& \multirow{2}{*}{$2 \numtmps \resampsamps\ f_1 / f_2$} \\
% Resampling contd...
$\,\,\,\,\,\,$ and sample rates $f_1 < f_2$ & \\
% Matrix multiply
multiply $M \times L$ real matrix by $L\times1$ real vector & $2 M L$ \\
%
\end{tabular}
\end{center}
\caption{Number of floating point operations per sample (multiplications and
divisions) required for a selection of signal processing operations used in
\textsc{lloid}.}
\label{table:flops}
\end{table}

\begin{comment}
The filter bank can be implemented using finite impulse response (\fir{}) filters,
which are just sliding window dot products.  If there are \numtmps\ templates
of length \tmpsamps\, and the data stream contains \hoftsamps\ samples, then
applying the filter bank requires $2 \numtmps \tmpsamps \hoftsamps$ operations.

More commonly, the matched filters are implemented using the \fft\ convolution.
This entails applying {\fft}s to blocks of \fftblock\ samples, with $\tmpsamps
\leq \fftblock$, each block overlapping the previous one by $ \fftblock -
\tmpsamps $ samples.  There are $\hoftsamps/(\fftblock-\tmpsamps)$ such blocks
required to filter \hoftsamps\ samples of data.  Modern implementations of the
Cooley-Tukey \fft, such as the ubiquitous \texttt{fftw}, require about $4 N \lg
N$ operations to evaluate a \textsc{dft} of size $N$~\cite{Johnson:2007p9654}.
%
%
\editorial{This is more commonly known as ``overlap-save''.  We should find
someone else's operation count and cite it.}  
%
%
A \fftblock\ sample cross-correlation consists of a forward \fft, a \fftblock
sample dot product, and an inverse \fft\, totaling $8 \fftblock \lg \fftblock +
2 D$ operations per block.  Per sample, this is $(8 \lg \fftblock + 2) / (1 -
\tmpsamps/\fftblock)$ operations.  As this expression indicates the number
of operations increases as the block size \fftblock\ approaches the
filter length \tmpsamps.
%
%
\editorial{Drew: Why don't we change this to an overlap of $m$ samples so we
can see what happens as we increase the overlap to reduce latency.}

The \fir\ filter implementation has the advantage that it has no intrinsic
latency, whereas the \fft\ convolution has latency of $\fftblock-\tmpsamps$. 
%
%
%For example, for a $1.4 -
%1.4 \, M_\odot$ template with duration $\sim 1 \, \mathrm{ks}$, the
%\textsc{fft} convolution has a latency $\geq 2 \, \mathrm{ks}$.  
%
However, the \fir\ filter implementation has the disadvantage of much greater
overhead per sample than the \fft\ convolution.  For a $1\,\mathrm{ks}$
template sampled at $4096\,\mathrm{Hz}$, the \textsc{fir} implementation
requires about about $\tmpsamps / 8 \lg 2 \tmpsamps = 2.2 \times 10^4$ times
more operations per sample than the \fft{} implementation.  We will now consider
the computational cost of the \fir\ filter implementation described in the
previous sections.

It is convenient to express the computational cost of the entire filtering
procedure in floating point operations per second {\flops}.  The cost will be
the sum of the cost of the \fir\ filtering for the orthogonal filter in each
time slice plus the cost of reconstructing the original waveforms with matrix
operations and resampling.  Using the formulas in table \ref{table:flops} we
arrive at
%
\begin{equation}
%\mathrm{Computational\, cost} 
\mathrm{C.C.} \propto 2 \sum_k^{\numslices} (\slicessamps + \numtmps) \numsvdtmps f_k \nonumber
	r \sum_{k, f_k \neq f_{\fmax}}^{\numslices} \numtmps \resampsamps f_k 
\label{eq:TDcost}
\end{equation}
%
where the first sum is the total cost for filtering and reconstructing the
orthogonal filters $u_{jk}$.  The second sum is the cost of resampling the
reconstructed time-slice outputs to the original sample rate {\fmax}.  It
assumes a \fir\ filter with \resampsamps\ sample points.  Note that the cost
for resampling only occurs for the downsampled time slices. The resampling cost
largely depends on {\resampsamps}.  The computational cost of \eqref{eq:TDcost}
is dominated by the highest frequency terms in the sum.  Comparing just the
highest frequency term of \eqref{eq:TDcost} with \eqref{eq:SNRTD} shows \order{
(\slicesamps{\mathrm{max}} + \numtmps) \svdtmps{\mathrm{max}} \fmax} \flops{}
versus \order{\numtmps \tmpsamps \fmax} {\flops}.  The full calculation for a
particular patch of parameter space is shown in table \ref{table:comp_cost}.
\hl{FIXME we need to actually say what went into this}

\editorial{From here down introduce an example, but don't go through the FFT
methods in lloid focus just on the comparison of the TD method without lloid
tricks, the FD method without lloid tricks, and the TD lloid method (no comp
detection statistic though), remember that the goal is near realtime detection
not computational cost improvements on FD methods}

\end{comment}

>>>>

We now examine the computational cost scaling of the approximate implementation
of a conventional time domain or frequency domain matched filter bank as compared
with \lloid{}.  For convenience, \ref{tab:recap} provides a review of the notation
that we will need in this section.
%
% symbols used in FLOPs calculations table
%
\Table{\label{tab:recap} Notation used to describe filters.}
\br
Symbol	& Definition \\
\mr
\numtmps		& The number of templates \\
\tmpsamps	& The number of samples per template \\
\numslices	& The number of time slices \\
\numsvdtmps	& The number of orthogonal templates in time slice $s$ \\
\slicessamps	& The number of samples in time slice $s$ \\
$f^s$		& The sample rate in time slice $s$ \\
$N^\shortdownarrow$ & The number of coefficients in the decimation filter \\
$N^\shortuparrow$ & The number of coefficients in the interpolation filter \\
\br
\end{tabular}
\end{indented}
\end{table}

\Table{\label{table:flops}Minimum computational cost of the \textsc{fir}, \textsc{fft} convolution, and \lloid\ methods for a bank of 657 templates sampled at 4096 Hz.}
\br
Method & \flops \\
\mr
\textsc{fir} & $2.4\times10^{13}$ \\
\textsc{fft} convolution & $2.6\times10^8$ \\
\lloid & $4.7\times10^8$ \\
\br
\end{tabular}
\end{indented}
\end{table}

\subsubsection{Conventional time domain method}

The conventional time domain method consists of a bank of \textsc{fir} filters, or sliding window dot products.  If there are $\numtmps$ templates, each $\tmpsamps$ samples in length, then each filter requires $M N$ multiplications and additions per sample, or $2 \numtmps \tmpsamps f^0$ floating point operations per second (\flops) at a sample rate $f^0$.

\subsubsection{Conventional frequency domain method}

The most common frequency domain method is known as the \emph{overlap-save} algorithm, described in \cite{numerical-recipes-chapter-13}.  It entails splitting the input into blocks of $D$ samples, $D > \tmpsamps$, each block overlapping the previous one by $D - \tmpsamps$ samples.  For each block, the algorithm computes the forward \fft\ of the data and the templates, multiplies them, and then computes the reverse \fft.

Modern implementations of the \fft, such as the ubiquitous \texttt{fftw}, require about $2 \fftblock \lg \fftblock$ operations to evaluate a real \textsc{dft} of size $\fftblock$~\cite{Johnson:2007p9654}.  Including the forward transform of the data and $M$ reverse transforms for all of the templates, the \fft\ costs $2 (\numtmps + 1) \fftblock \lg \fftblock$ operations per block.  The multiplication of the transforms adds a further $2 \numtmps \fftblock$ operations per block.  Since each block produces $\fftblock - \tmpsamps$ usable samples of output, the overlap-save method requires
$$
f^0 \cdot \frac{2 (\numtmps + 1) \lg \fftblock + 2 \numtmps}{1 - \tmpsamps/\fftblock} \; \mathrm{\flops}.
$$

\subsubsection{\lloid\ method}

For time slices $s$, the \lloid\ method requires $2 \slicessamps \numsvdtmps f^s$ for evaluating the orthogonal filters, $2 \numtmps \numsvdtmps f^s$ \flops\ for the linear transformation from the $\numsvdtmps$ basis templates to the $\numtmps$ time-sliced templates, and $\numtmps f^s$ \flops to accumulate the resultant partial \SNR\ stream.

The computational cost of the decimation of the detector data and the interpolation of the partial \SNR\ streams is a little bit more subtle.  Decimation is achieved by applying an \fir\ antialiasing filter and then downsampling, or deleting samples in order to reduce the sample rate from $f^{s-1}$ to $f^s$.  Naively, the antialiasing filter with $N^\shortdownarrow$ coefficients should demand $2 N^\shortdownarrow f^{s-1}$ \flops.  However, it is only necessary to evaluate the antialiasing filter for the fraction $f^s / f^{s-1}$ of the samples that will not be deleted.  Consequently, an efficient decimator that requires only $2 N^\shortdownarrow f^{s-1} \cdot \left( f^s / f^{s-1} \right) = 2 N^\shortdownarrow f^s$ \flops\ is possible.

The story is similar for the interpolation filters used to change the sample rates of the partial \textsc{snr} streams.  Interpolation of a data stream from a sample rate $f^s$ to $f^{s-1}$ consists of inserting zeros between the samples of the original stream, and then applying a lowpass filter with $N^\shortdownarrow$.  The lowpass filter requires $2 M N^\shortdownarrow f^{s-1}$ \flops.  However, by taking advantage of the fact that by construction a fraction $f^{s-1}/f$ of the samples are zero, it is possible to build an efficient interpolator that requires only $M N^\shortuparrow f^{s-1} \cdot \left( f^s / f^{s-1} \right) = 2 M N^\shortuparrow f^s$ \flops.

Taking into account the decimation of the detector data, the orthogonal \fir\ filters, the reconstruction of the time sliced templates, the interpolation of \SNR\ from previous time slices, and the accumulation of \SNR, in total \lloid\ requires
$$
\sum_{\mathclap{s=0}}^{\mathclap{S-1}} \left( 2 \slicessamps \numsvdtmps + 2 \numtmps \numsvdtmps + \numtmps \right) f^s + 2\sum_{\mathclap{f \in \{f^s\}}} \left( N^\shortdownarrow + \numtmps N^\shortuparrow \right) f \textrm{ \flops.}
$$
The second sum is carried out over the set of sample rates of all of the time slices rather than over the time slices themselves because some time slices may have the same sample rates.
